Concept of a Double Tilted Rowland Spectrograph for X-rays

### 2024-08-20 papers/Doubletiltconcept/referee_reply_1.txt

We thank the referee for their time spent reviewing the manuscript. The quality of the feedback expressed in the review makes it apparent that the referee is an expert in the Rowland geometry. However, as we note in our response below, we disagree that the general principles of the geometry are as widely known by the broader astronomical community as claimed by the referee. Moreover, none of the prior discussion of the DTRS concept appears in peer-reviewed publications. This paper is intended to address these shortcomings of the literature.

The referee has made some specific suggestion with figures to combine or remove and we address those below (note that this means that figures in the new manuscript are renumbered and might not match the figure numbers in the previous referee's comments any more). We highlighted new or changed text in bolt font for the re-submission, but we also shortened several of the discussions in the paper, consistent with the referee’s suggestion to focus on information not presented elsewhere. Given the difficulty of reading through the substantial amount of strike-through figures and strike-through text, we opted to fully remove (instead of strike-through) text that we no longer need.

Below, our specific responses to the referee's comments are keyed in between the referee's comments.

Reviewer:
The paper gives a full description of the 'concept of a Double Tilted Rowland Spectrograph for X-rays' but the level of new information is rather limited given the extensive literature by the authors and the reference to FUSE (moose 2000) and many others .

>>> We want to point out that the FUSE design (Moose 2000) is similar in some aspects, but very different on a few key points:
- FUSE uses reflective gratings, not transmission gratings.
- In FUSE, the gratings are illuminated at almost normal incidence and thus entrance aperture and detectors are on the same side of the Rowland circle, not on opposite sides as in our design.
- FUSE does not detect the zeroth orders.
- FUSE does not place all diffracted spectra next to each other on the same read-out array.

For publishing it in ApJ my recommendation is to limit the paper to the really new information and not describe the details (and even repeat them in the paper itself). From my understanding the really new information is probably:

- The use of an offset in the cross dispersion direction
- The discussion about alignment accuracies but, these are only described qualitatively whereas the new information (using the tools) is probably the quantitative information
- Correctly one of the identified drawback is the background (in section 4.2.2). Quantifying under which conditions DTRS or a more classical Rowland angle design is better would be a really new contribution
- Various aspects are not discussed in the paper including order separation and mirror scattering in the cross dispersion direction. Probably this is nicely included on the available tools but how this impacts the performance is not quantified

Hence I would recommend to ask the authors to rewrite the paper significantly, limiting it to the real new information compared to previous articles.

>>> We believe that the comments above are partially related to a misunderstanding about what has been published in "previous articles". It is true that we have written about the concept of the Double Tilted Rowland Spectrograph (DTRS) before specifically for the Arcus mission. However, that work so far has only appeared in conference proceedings (Proceedings of the SPIE), not in refereed journals. We believe that a prior publication in a conference proceeding should not preclude publication of a proper article. Those previous proceedings are also specific to Arcus: They use the Arcus focal length, the specific size in mm for the gratings that Arcus is proposed to have etc. On the one hand, that allows us to be very specific about the resolving power and effective area of the instrument in those conference proceedings, but does not include a general discussion of underlying CONCEPT in general. In this article, the discussion is necessarily more qualitative (because it is not specific to a specific number for focal length, grating size etc.), but more general, so we believe that it adds some value even for those readers who are familiar with with our previous proceedings; we thus regard content as "new" even if it has been described specifically in the Arcus proceedings before.
We have added sentences to the introduction to make this goal and the quality of prior work explicit.
We have also gone through the text to remove duplication and improve the flow by shortening, in particular in section 2, which describes the background for our work. In total, the re-submitted work is about 1.5 pages shorter than our original submission.

For previous papers you can also look also the following papers although the paper by Moose describes most of the technology.

Varied Line-Space Gratings: Past, Present and Future by Michael C. Hettrick, Proc. Soc. Photo-Opt. Instr. Eng. vol. 560, pp. 96-108 (1986)
http://dx.doi.org/10.1117/12.949620

>>> We have added this reference to the section about "dispersive elements" (sect 2.2).

Space Sci Rev (2010) 157: 15-24 DOI 10.1007/s11214-010-9719-y, X-ray Diffraction Gratings for Astrophysics by Frits Paerels

>>> Given the length of the paper (which the referee suggested to shorten significantly), we had chosen to not discuss those review articles. However, we agree that this is a useful article and added a reference to this paper.

Draft version June 22, 2022 Typeset using LATEX default style in AASTeX631 X-ray performance of critical-angle transmission grating prototypes for the Arcus mission by Heillman

>>> This article was already cited and discussed in the version we initially submitted. That work discusses the details of how the gratings are made and how efficient they are, but not how they are used in a Double-Tilted Rowland Spectrograph.

All figures: add coordinates to the plots (X, Y, Z (is the direction of the X-rays?)

>>> Fig 1 and 2 already had x,y,z. For the 3D plots actually adding a floating coordinate system that stays visible and renders well in the webbrowser is surprisingly difficult and required significant effort, but it has now been added to the 3D renderings. In Fig 9, we just describe the relation to the global X,Y,Z in the figure caption, since the dispersion coordinate is read along the curved Rowland circle. In our layout, not even at the focal points does this exactly coincide with the x axis of the coordinate system.

Datailed comments (apart of the length of the paper):

Section 2.1 end: figure 1 is well known and does not need to be illustrated and can be summarized shortly.

>>> When we presented this or similar work in talks, we've found that even people who were involved in the design of Chandra are surprised when we point out the LETG and HETG would have higher resolving power, had Chandra chosen to sub-aperture. It thus seems that the community to which this is "well known" is limited to instrument designers. Since ApJ is a general purpose astronomy journal we've chosen to include this as a figure to make it clear. We admit that this figure does not contain "new information" but merely illustrates previous knowledge, but given how central this is this to the rest of the article, we suggest that it is worthwhile to show in a figure and have thus left it in this version - if this argument does not convince the referee, we will remove it in the next resubmission.

Section 2.2: a correct statement about 'larger gratings (line 7) but this is not the topic of this paper)

>>> Grating size is not the main topic of this paper, but the main motivation for Heilmann et al. to introduce a tilted Rowland torus to transmission gratings was to reduce the average deviation between grating and Rowland torus. For small gratings, that distance is always small, so it seems relevant to at least bring up the topic of grating sizes.

Section 3 Figure 3: in figure 2 the suggestion is given that the mirror covers a circular sub-aperture of the full mirror plane while in the ARCUS paper this is much clearer described (two 'rectangular' subsections. This would also make figure 4 more understandable.

>>> We like the referee's suggestion (in a point below) to also add the aperture to the figure that shows the distribution of photons on the focal plane (currently Fig 5). That clearly shows the area covered by mirror+gratings.
In Fig 2, we have shown a tilted, but not sub-apertured scenario. In our setup for this scenario, the mirror does cover the full circle, not a rectangular subsection.
We also want to note that, in our previous version, we had chosen to use wedges, not rectangles for layout of the gratings in the sub-apertures. For the purpose of this article though the exact shape does not matter - our simulations are meant to illustrate the concept, not to match any particular mission. However, since our use of wedges seems to be unclear, we have redone this now with rectangles for the grating layout to better match the shape of the green boxes we use for the mirrors in our visualization. (Arcus uses something in between that is not quite a rectangle, but is wider at the tip than at the bottom.)

Section 3.2 Figure 4: what is the added value compared to fig 2

>>> In the electronic version, Fig 4 is interactive and Fig 2 is not. For the static version of the article, Fig 4 shows four views including a view of the focal plane, and a zoom-in on the gratings, which is not visible in Fig 2.

Section 3.2 just below figure 4: The use of tilted Rowland circles is used already for a long time in space missions (see RGS on XMM-Newton). The new step introduced by Heilman is that this can also be achieved by Critical Angle Transmission gratings. In this sense using tilted Rowland cicles is a well known concept.

>>> We re-worded that sentence.

Frig 5: blue and black colors are not easy to distinguish. Also the figure becomes more clear if at the rich side a panel is shown with the mirror aperture used for each of the sub-panels.

>>> There were no black points in Fig 5, maybe the referee meant dark purple and blue? We understand the point, but there are are only so many colors we can choose from, also taking into account common combinations to avoid for accessibility (e.g. red/green blindness). We've re-done the figure with different colors. We thank the referee for the suggestion to add the mirror aperture to the figure.

In the same paragraph (5 lines lower) high diffraction efficiencies are mentioned. Omn the various simulatiopns/descriptions this is further neglected. It wopuld be good to quantify this.

>>> We added a number and a reference.

Equation 3: I do not believe anywhere the y component is used (but probably it is included as it is part of the software tools (but then y not equals zero).

>>> We want to be explicit here that the focal point is indeed in the plane that we show in the sketches - the sketches are not a projection onto the z=0 plane, but a cut at the z=0 plane and the focal point is in that plane. (With the addition of the x,y,z vectors to all images, we standarized the coordinate system so that z is now the cross-dispersion direction, so whatever was previously called y is now z.)

Section 3.3: Figure 6 and 7 are another set of similar figures, reducing it to fig 7 should be considered as in fig 2 you describe the basic principle and in fig 7 you give the final concept.

>>> Going from one Rowland Torus to two is a pretty big conceptual step. We have explained the development of the double-tilted Rowland torus in several talks with different audiences (both observing astronomers and instrument builders) and always found that it is conceptually challenging, thus we want to make sure that there is enough detail in the article to explain to the reader how we get there, not just "what the answer is". Thus, our attempt to break it in four steps:
1. Chandra-like (and thus well known to many in the community) - old Fig 4
2. tilted torus (old Fig 6)
3. two tilted tori (old Fig 7)
4. two tilted tori, with the two parts of the two tori offset from each other (old Fig 8).

However, we also understand the referee's concern that the article is long and, for the expert reader, contains too much material that is repeated. So, in order to reduce the number of figures, we've removed the old Fig 7 since Fig 7 and 8 look much more similar than Fig 6 and 7.
(Note however, that by reordering material and removing figures, the figure numbers have changed in the resubmitted version.)

Section 3.4.1 is probably the most relevant extension of earlier papers although I did not check if this is not already in the published ARCUS proposal

>>> We want to be explicit that there is no such thing as a "published ARCUS proposal". Like all proposals, the Arcus proposal submitted to NASA is strictly confidential; even if the referee has access to this proposal in some capacity, it is not relevant to the review of this article.
The goal of this article is to describe the general concept in the refereed literature, as opposed to previous conference proceedings where we always wrote about one specific mission concept.

Section 4.1.1: The advantage of subaperturing is presented at the cost of the technical compelkxity achieving a better PSF. As such it is true but one limits the usable open area provided by a satellite (with a circular shaoe). Alternatively one populate the unsused open area at the cost of an ectra detector to collect the dispersed X-rays in a 90 degree rotated system

>>> We have looked at extra detectors before, but one goal of the DTRS design is to avoid this. We added a short discussion of this alternative.
The question of what is a better use of the open area depends, as the referee points out, on the shape of the opening and the shape of the mirror. For example, in Arcus, the open area is constrained by the shape of the boom (square) while in Chandra the open area is a circle. For Arcus or Athena where the mirror is made out of stacks of SPOs the "mirror shape" can be arranged in different ways for a sub-aperture (e.g. a square or a wedge), while mirrors using full shells like Chandra cover a circle. Thus, there can be no general answer - it depends on the specific mission design and is not an advantage of disadvantage of the DTRS as a general concept.
We discussed a setup like the referee suggested (with a 90 degree rotated system) in a previous paper focussed on the Lynx mission. We have added a reference to that paper.

Figure 9: it could be good to identify which parts of the recorded data is to to the dispersion of the photons perpendicular to the main dispersion direction. How does this affect the dispersed spectram?

>>> While this is an interesting discussion, we believe that it is beyond the scope of this article. The referee already pointed out that the length of the article is a concern, and thus we intentionally left out the problem of order-sorting and a quantitative analysis of cross-dispersion. Both depend sensitively on the specific design and this this is hard to generalize without discussing a specific mission concept, which is not in scope for ApJ instrumentation papers. For example, the fraction of photons that are cross-dispersed depends almost linearly on the spacing of cross-dispersion structure and the dimensions of the support structure. Prototype gratings for Arcus use bars with 0.9 mu width for the support structures in cross-dispersion direction, but e.g. for the gratings Lynx the bars would have been thinner and thus the fraction of photons that are cross-dispersed would be lower; one could also test if larger spacing of the grating support structures in the cross-dispersion direction are sufficient for structural stability and reduce the cross-dispersion that way. So it really depends on the specific gratings and cannot be answered in general.

Fig 10: contains essentially the same as figure 9, I would reduce this to a single figure

>>> We originally included Fig 10 because some details are better seen in the larger scale compared to Fig 9, but the referee is correct: The same data is shown and thus we removed Fig 10 and the associated text to reduce the length of the article.

### 2024-08-20 papers/Doubletiltconcept/doubletiltconcept.tex
second revision of Double tilted concept paper

\author[0000-0003-4243-2840]{Hans Moritz G{\"u}nther}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\author[0000-0002-9184-4561]{Casey DeRoo}
\affiliation{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
\author[0000-0001-9980-5295]{Ralf K. Heilmann}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\affil{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\author[0000-0002-6747-9648]{Edward Hertz}
\affiliation{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}

@@

%% From the front matter, we move on to the body of the paper.
%% Sections are demarcated by \section and \subsection, respectively.
%% Observe the use of the LaTeX \label
%% command after the \subsection to give a symbolic KEY to the
%% subsection for cross-referencing in a \ref command.
%% You can use LaTeX's \ref and \label commands to keep track of
%% cross-references to sections, equations, tables, and figures.
%% That way, if you change the order of any elements, LaTeX will
%% automatically renumber them.
%%
%% We recommend that authors also use the natbib \citep
%% and \citet commands to identify citations.  The citations are
%% tied to the reference list via symbolic KEYs. The KEY corresponds
%% to the KEY in the \bibitem in the reference list below.

@@

The two largest X-ray observatories in history both field diffraction gratings: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
Unlike microcalorimeters, diffraction gratings require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal. For a given diffraction order, gratings typically deliver a fixed wavelength resolution $\Delta \lambda$, so they perform better than microcalorimeters at longer wavelengths/lower energies. One common layout for a diffraction spectrometer is the Rowland torus \citep{Beuermann:78}. In the plane defined by the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other. This circle is commonly called the ``Rowland circle''. In three dimensions, the surface that defines the best grating positions has the shape of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors. The general concept could be applied to a large range of focal lengths, grating constants, and other hardware properties. Thus, we discuss advantages and challenges of this layout in general. However, we note that Arcus \citep{2023SPIE12678E..0ES}, an X-ray spectroscopy mission proposed for NASA MIDEX (medium explorer) and Probe calls, applies this layout. Specific details of the  Arcus performance, such as predicted effective area and resolving power are shown in \citet{2018SPIE10699E..6FG,2023SPIE12678E..1DG}, but the general concept of a double tilted Rowland spectrograph (DTRS) with its advantages and drawbacks has not been discussed -- a void we attempt to fill with this article.

The structure of this article is as follows: First, we discuss the properties of the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, we illustrate the optical layout, modifying the design of a traditional single-channel Rowland torus spectrograph to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}). We end with a short summary in section \ref{sect:summary}.

@@

\subsection{Mirror characteristics and sub-aperturing}
The concept of a DTRS can be implemented with different mirror technologies. In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence. The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for efficient reflection. The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases. Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements. In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape. If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter. In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C,2010SPIE.7732E..1JH}. This is illustrated in Fig.~\ref{fig:subaperture}.

\begin{figure*}
\plotone{subaperture.pdf}
\caption{
    Illustration of the concept of sub-aperturing. All panels show a view from the universe looking down onto the telescope. \emph{left}: Each dot is a photon shown as it passes through a diffraction grating. Photons are colored according to their polar angle in this panel. \emph{middle and right:} Photons detected in the zeroth and first diffraction order on the detector, respectively. The dispersion direction is from left to right ($y$ coordinate), the distance from the focal point at $(y, z) = (0, 0)$ is shown. Each photon is shown in the same color as in the left panel. Photons passing mirror and grating close to $z=0$ (dark purple) are spread out along the dispersion direction ($y$ coordinate); photons passing mirror and gratings close to $y=0$ are more concentrated along the dispersion direction. Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction. Sub-aperturing thus reduces the PSF in the dispersion direction, which increases the spectral resolving power.}
\label{fig:subaperture}
\end{figure*}

\subsection{Dispersive elements}
In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus, ideally tangentially for transmission gratings. They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
\begin{equation}
n \lambda = d \sin \theta \label{eqn:diffraction}
\end{equation}
where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the incoming light and the normal to the grating. (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings and does not vary within a grating. While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly. Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus. \citet{2020SPIE11444E..88G} explored bending the gratings in one dimension or varying the distance between grating bars over a grating to compensate for the deviations from the ideal shape in simulations. \citet{2019SPIE11118E..11G} shows that CAT gratings can be bent to the required radii without degrading their performance.

\subsection{Detectors}
The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other. Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectrum, it is beneficial to keep the number of elements low. In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive in 2D to separate individual channels.  Because it covers a larger area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design. The detector should follow the shape of the Rowland circle. This can be achieved by using curved detectors or by tiling smaller flat detector elements where each element is tangential to the Rowland circle.

The gratings spread out the signal over a large physical region on the detector. Thus, the counts rates per pixel are much lower than they would be in an imaging observation. Detectors that might be at risk of pile-up in direct observations can still be used for the dispersed signal. On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) make up a larger fraction of the extracted signal than in direct imaging observations. Similarly, fast timing resolution is typically not essential, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway. Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location, and to separate grating orders where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~nm photons, close to the O~{\sc viii} Lyman $\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).

@@

\section{Optical Layout}
\label{sect:opticallayout}
In this section, we describe the optical layout for a DTRS. We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS. We then show how this design can be modified step-by-step to accommodate blazed gratings and what considerations are important when choosing the specific parameters for a design. We illustrate the concept with ray-trace simulations. Those simulations are performed with the marxs code \citep{2017AJ....154..243G}. However, the setup is chosen to illustrate the concept of a DTRS, not to represent any particular instrument. For example, the position of the diffractive gratings shown leaves valuable mirror area uncovered and is done without considering the size and location of support structures that the gratings can be mounted on. The grating constant $d$ is chosen to be smaller than feasible for a real instrument. Static images of the 3D ray-traces are embedded in the article, but we encourage the reader explore the interactive 3D models embedded in the electronic version of the publication.

The points $\vec p$ on the surface of a torus can be parameterized by two angles $\varphi$ and $\theta$.
\begin{equation}
\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_y \sin \theta + \vec e_R(\varphi) \cos \theta)
\end{equation}
where $\vec c$ points to the center of the torus. The axis of symmetry is given by a unit vector $\vec e_y$. We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_z \sin\varphi$. $R \vec e_R$ points from the center of the torus to the center of a circle with radius $r$. Sweeping that circle around the symmetry axis forms the torus. If $r < R$, the torus has an inner hole like a doughnut, and if $r > R$ the inner parts of the torus overlap with each other (``spindle torus'').

>>>>

\author[0000-0003-4243-2840]{Hans Moritz G{\"u}nther}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\author[0000-0002-9184-4561]{Casey T. DeRoo}
\affiliation{Dept. of Physics \& Astronomy, University of Iowa, Iowa City, IA 52242, USA}
\author[0000-0001-9980-5295]{Ralf K. Heilmann}
\affiliation{MIT Kavli Institute for Astrophysics and Space Research, 77 Massachusetts Avenue, Cambridge, MA 02139, USA}
\affil{Space Nanotechnology Laboratory, MIT Kavli Institute for Astrophysics and Space Research, Cambridge, MA 02139, USA}
\author[0000-0002-6747-9648]{Edward Hertz}
\affiliation{Center for Astrophysics, Harvard-Smithsonian Astrophysical Observatory, Cambridge, MA 02138, USA}

@@

@@

The two largest X-ray observatories in history both field diffraction gratings \citep[see][for a review]{2010SSRv..157...15P}: The Reflection Grating Spectrometer (RGS) on XMM-Newton \citep{2001A&A...365L...7D} and both the High-Energy Transmission Grating  \citep[HETG,][]{2005PASP..117.1144C} and the Low-Energy Transmission Grating \citep[LETG,][]{1997SPIE.3113..172P} on Chandra.
Unlike microcalorimeters, grating spectrographs require an additional dispersive element (a transmission or a reflection grating) in the optical path of the instrument and dedicated detectors to capture the dispersed signal. For a given diffraction order, gratings typically deliver a fixed wavelength resolution $\Delta \lambda$, so they perform better than microcalorimeters at longer wavelengths/lower energies. One common layout for a diffraction spectrometer is the Rowland torus \citep{Beuermann:78}. In the plane defined by the optical axis and the dispersion direction, the gratings are located at one side of a circle, the detectors on the other. This circle is commonly called the ``Rowland circle''. In three dimensions, the surface that defines the best grating positions has the shape of a torus, thus the name ``Rowland torus'' for this geometry (see Sect.~\ref{sect:onetorus} for a figure).

In this article, we present an optical layout for an instrument that stacks two or four different channels with distinct optical axes in such a way that each channel follows the Rowland torus prescription, yet at the same time all channels are imaged onto a common set of detectors.

This concept has been developed and described in the context of the Arcus mission, which was originally proposed as a NASA Medium Explorer \citep{2019SPIE11118E..0WS} and  has since been proposed to the NASA Astrophysics Probe call \citep{2023SPIE12678E..0ES}. However, the DTRS concept is applicable to mission architectures beyond that of Arcus, and could be applied to a broad range of focal lengths, grating parameters, etc. This paper aims to discuss the advantages and challenges of this layout from a general perspective. We note that this discussion will be necessarily qualitative (rather than quantitative) as performance metrics such as resolving power and effective area depend on the design parameters that will be specific to a given instrument (e.g., focal length, efficiency).

The structure of this article is as follows: First, we discuss the properties of the optical elements that are most influential for the design of a DTRS, namely mirror properties, gratings, and detectors in Sect.~\ref{sect:elements}. Then, we illustrate the optical layout, starting by modifying the well-known single-channel Rowland torus spectrograph design step-by-step to a DTRS. In section~\ref{sect:discussion}, we discuss advantages and disadvantages of our DTRS concept in general and for the example of Arcus in particular (section~\ref{sect:applications}). We end with a short summary in section \ref{sect:summary}.

@@

\subsection{Mirror characteristics and sub-aperturing}
The concept of a DTRS can be implemented with different mirror technologies. In X-rays, the mirror assembly will typically consist of nested (full or partial) shells that reflect X-rays in grazing incidence. The higher the energy of the X-ray, the smaller the angle between the mirror surface and the incoming ray has to be to allow for efficient reflection. The size of the collecting area is to first order proportional to the number of photons that the spectrograph can detect, and the spectral resolving power $R=\lambda / \Delta \lambda$ increases when the mirror point-spread-function (PSF) decreases. Depending on the type of mirror chosen, the PSF can be dominated by the alignment error between mirror shells, in particular for mirror assemblies that consist of hundreds of elements. In this case, X-rays are offset from the focus point both in the plane of reflection and perpendicular to it.
On the other hand, mirrors can also have a figure error, which is the deviation of the mirror surface from the ideal shape. If the PSF is dominated by the figure error or scattering by microroughness or particulates, then scattering angles in the plane of reflection are typically larger than the out-of-plane scatter. In this case, sampling a limited azimuthal span of the aperture -- subaperturing -- can improve the spectral resolving power \citep{1987ApOpt..26.2915C,2010SPIE.7732E..1JH}. This is illustrated in Fig.~\ref{fig:subaperture}.

\begin{figure*}
\plotone{subaperture.pdf}
\caption{
    Illustration of the concept of sub-aperturing. All panels show a view from the universe looking down onto the telescope. \emph{left}: Each dot is a photon shown as it passes through a diffraction grating. Photons are colored according to their polar angle in this panel. \emph{middle and right:} Photons detected in the zeroth and first diffraction order on the detector, respectively. The dispersion direction is from left to right ($y$ coordinate), the distance from the focal point at $(y, z) = (0, 0)$ is shown. Each photon is shown in the same color as in the left panel. Photons passing mirror and grating close to $z=0$ (dark purple) are spread out along the dispersion direction ($y$ coordinate); photons passing mirror and gratings close to $y=0$ are more concentrated along the dispersion direction. Sub-aperturing means that the mirror and gratings do not cover a full circle, but only regions marked in yellow and green in the left panel, so that the resulting photon distributions on the detector are narrower in dispersion direction, which increases the spectral resolving power.}
\label{fig:subaperture}
\end{figure*}

\subsection{Dispersive elements}
In a Rowland torus geometry, diffraction gratings are placed on the surface of the Rowland torus, ideally tangentially for transmission gratings. They diffract light according to the grating equation for photons arriving perpendicular to the grating surface:
\begin{equation}
n \lambda = d \sin \theta \label{eqn:diffraction}
\end{equation}
where $\lambda$ is the wavelength of the light, $n$ is the order of the diffraction, $d$ is the grating constant, and $\theta$ is the angle between the outgoing light and the normal to the grating. (The equation can be rewritten for rays hitting the grating at arbitrary angles, but for the purpose of this conceptual discussion this is not needed.)
Typically, these transmission gratings are flat and only a few cm$^2$ in size; the spacing between gratings bars $d$ is the same for all gratings.
% and does not vary within a grating.
While a flat grating necessarily deviates from the curved surface of a torus, in today's instruments those deviations are small enough that they do not impact the resolving power of the spectrograph significantly. Larger gratings would increase the effective area of the instrument because less space is needed for frames and other mounting structures that hold the gratings in place, but they would also deviate more from the surface of the torus thus reducing $R$.
To prevent a loss of $R$ with larger gratings one can use gratings where $d$ varies across the grating \citep{1986SPIE..560...96H,2020SPIE11444E..88G} or bend the gratings \citep{2020SPIE11444E..88G}. \citet{2019SPIE11118E..11G} shows that gratings can be bent to the required radii without degrading their performance.

\subsection{Detectors}
%The dispersed spectrum is spread out far in the dispersion direction. Thus, the ideal detector is much longer in one direction than in the other.
%Individual detector elements (e.g.\ CCDs) can be tiled, but to avoid chip gaps where photons are lost and gaps appear in the extracted spectrum, it is beneficial to keep the number of elements low.
In our DTRS design several spectral traces are positioned in parallel (section~\ref{sect:opticallayout}), so the detector has to be position sensitive to separate individual channels.  Because it covers a larger area than in direct imaging observations, a detector with moderate cost and power requirements per surface area is well-suited for this design. The detector should follow the shape of the Rowland circle. This can be achieved by using curved detectors or by tiling smaller flat detector elements. % where each element is tangential to the Rowland circle.

The gratings spread out the signal over a large region on the detector. Thus, the count rates per pixel are lower than they would be in an imaging observation. Detectors that might be at risk of pile-up in direct observations can be used for the dispersed signal. On the other hand, the larger regions needed to extract the signal imply that background events (internal background and diffuse astrophysical emission) make up a larger fraction of the extracted signal than in direct imaging observations.
%Similarly, fast timing resolution is typically not essential, because the lower signal means that the detected photons need to be binned up into longer time bins for analysis anyway.
Intrinsic energy resolution is required to both suppress the background by filtering out events that are incompatible with wavelengths of the dispersed signal at any particular location, and to separate grating orders where photons of different orders are diffracted to the same position (e.g.\ the sixth order for the O~{\sc vii} forbidden line of the He-like triplet at 2.21~nm and the seventh order of 1.94~nm photons, close to the O~{\sc viii} Lyman $\alpha$ line end up at almost the same $\theta$ according to equation~\ref{eqn:diffraction}).

@@

\section{Optical Layout}
\label{sect:opticallayout}
In this section, we describe the optical layout for a DTRS. We start from a single central Rowland torus as in Chandra/HETGS and Chandra/LETGS. We then show how this design can be modified step-by-step.
The general concept is shown using ray-traces with the marxs code \citep{2017AJ....154..243G} using a simple setup for illustration.
For example, the position of the diffractive gratings shown leaves valuable mirror area uncovered and is done without considering the size and location of support structures that the gratings can be mounted on. The grating constant $d$ is chosen to be smaller than feasible for a real instrument.

The points $\vec p$ on the surface of a torus can be parameterized by two angles $\varphi$ and $\theta$.
\begin{equation}
\vec p(\varphi, \theta) = \vec c + R \; \vec e_R(\varphi) + r \; (\vec e_z \sin \theta + \vec e_R(\varphi) \cos \theta)
\end{equation}
where $\vec c$ points to the center of the torus. The axis of symmetry is given by a unit vector $\vec e_z$. We define a vector $\vec e_R(\varphi) = \vec e_x \cos\varphi + \vec e_y \sin\varphi$. $R \vec e_R$ points from the center of the torus to the center of a circle with radius $r$. Sweeping that circle around the symmetry axis forms the torus. If $r < R$, the torus has an inner hole like a doughnut, and if $r > R$ the inner parts of the torus overlap with each other (``spindle torus'').

Figure~\ref{fig:sketch_label} shows sketches of the geometry in the plane spanned by the optical axis and the symmetry axis of the torus. Details will be discussed in the following subsections.
\begin{figure*}
    \includegraphics[width=0.5\textwidth]{sketch_label}
    \includegraphics[width=0.5\textwidth]{raytrace_labelled}
    \caption{\emph{left:} Sketch of a tilted Rowland torus geometry shown in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left. The location of the mirror is shown as blue box. The optical axis (solid blue line) is offset from the center of the Rowland circle (blue square). The symmetry axis of the torus  is marked as blue long-dashed line. \textbf{The torus is formed by rotating the blue circle around this line. To help visualize the 3D position of the torus blue dotted lines are shown connecting the two circles.} $\alpha$ is the angle between the optical axis and the line from the point where the optical axis intersects the Rowland torus (A) to the center of the Rowland circle. $\beta$ is the angle from optical axis to the line connecting A to the ``hinge'' -- the other point in this plane where symmetry axis of the torus and the Rowland circle intersect.
    \emph{right:} Visualization of a ray-trace of the same geometry. Colored lines show the path of individual rays starting on the left. Rays detected in the zeroth order are shown in green, blue and purple are higher orders. Optical axis and the axis of the Rowland torus are indicated in blue and the location of detectors in orange, matching the sketch in the left panel.
        }
    \label{fig:sketch_label}
\end{figure*}
\begin{figure*}
    \plotone{sketch}
    \caption{\emph{left:} Sketch of the on-axis Rowland geometry in the plane of the optical axis and the symmetry axis of the torus. Photons arrive from the left.
    \emph{center:} Tilted Rowland torus. The optical axis of the blue channel is offset from the center of the Rowland circle, but intersects the symmetry axis of the torus in the focal point $F$.
    \emph{right:} Layout with two Rowland tori, tilted in opposite direction. In the plane spanned by both optical axes (which are parallel) and the symmetry axis of the tori, the two tori overlap in exactly one circle, the Rowland circle. See figure~\ref{fig:sketch_label} for a more detailed explanation.
        }
    \label{fig:sketch}
\end{figure*}
